
# Normal distribution

This exposition is based on slides and text from Diez et al. 2015, OpenIntro Statistics (3rd Edition).


```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(magrittr)
library(cowplot)
```


## Basic about normal distributions

```{r,echo=FALSE, fig.width = 6, fig.height = 3.5}
npts <- 500
std.norm <- data_frame(x = seq(-4, 4, length.out = npts),
                       density = dnorm(x))

std.norm.plot <- 
  ggplot(std.norm, aes(x, density)) +
  geom_line() + 
  scale_x_continuous(breaks=c(-3, -2, -1, 0, 1, 2, 3),
                     labels=c("μ-3σ", "μ-2σ", "μ-1σ", "μ", "μ+1σ", "μ+2σ",  "μ+3σ"))+
  labs(y = "Density") +
  theme_classic() +
  theme(axis.title.x=element_blank()) 

std.norm.plot
```


- Unimodal 
- Symmetric
- Described by two parameters -- $\mu$ (mean) and $\sigma$ (standard deviation)
  
    - If a variable $X$ is apprximately normally distributed with mean, $\mu$, and standad deviation, $\sigma$, we write: $X \sim N(\mu,\sigma)$


## Normal distribution, probability density function

The probability density function for a normal distribution $N(\mu,\sigma)$ is described by the following equation:

\[
f(x|\mu,\sigma) = \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}
\]

```{r, echo=FALSE, fig.width = 6, fig.height = 3.5}
x.pt <- -0.5

std.norm.plot +
  geom_segment(x = x.pt, y = -1, xend = x.pt, yend = dnorm(x.pt),
               linetype = "dashed") + 
  geom_point(x = x.pt, y = dnorm(x.pt), color='red') +
  geom_text(x = x.pt-0.15, y = 0, label = "x") +
  geom_text(x = x.pt-0.25, y = dnorm(x.pt), label = "f(x)") +
  theme_classic() +
  theme(axis.title.x=element_blank()) 
  
```

- In R, the function `dnorm(x, mu, sigma)` calculates the probability density of $N(\mu,\sigma)$ at the point $x$


## Visualizing normal distributions 

- Different normal distributions look alike when plotted on their own scales

```{r, echo=FALSE, fig.width = 9, fig.height = 3.5}
norm.19.4 <- data_frame(x = seq(4, 34, length.out = npts),
                       density = dnorm(x, 19, 4))

std.norm.plot.labels <- 
  ggplot(std.norm, aes(x, density)) +
  geom_line() + 
  labs(y = "Density", title = "N(0,1)") +
  theme_classic()

norm.19.4.plot <- 
  ggplot(norm.19.4, aes(x, density)) +
  geom_line() + 
  labs(y = "Density", title = "N(19,4)") + 
  theme_classic()

plot_grid(std.norm.plot.labels, norm.19.4.plot, labels="AUTO")
```

- Must plot normals on a common scale to see the differences

```{r, echo=FALSE, fig.width = 9, fig.height = 3.5}
combined.norms <- 
  bind_rows(mutate(std.norm, distn="N(0,1)"),
            mutate(norm.19.4, distn="N(19,4)"))

ggplot(combined.norms, aes(x, density, color=distn)) +
  geom_line() + 
  labs(y = "Density", title = "N(0,1) and N(19,4) in common scale") + 
  theme_classic()
```


## Comparing values from different normal distributions


Q: SAT scores are approximately normally distributed with a mean of 1060 and a standard deviation of 195. ACT  scores are approximately normal with a mean of 20.9 and a standard deviation of 5.6. A college admissions officer wants to determine which of the two applicants scored better on their standardized test with respect to the other test takers: Malaika, who earned an 1350 on her SAT, or Jim, who scored a 28 on his ACT?

```{r, echo=FALSE, fig.width = 9, fig.height = 3.5}
SAT.norm <- data_frame(x = seq(480, 1600, length.out = npts),
                       density = dnorm(x, 1060, 195))

ACT.norm <- data_frame(x = seq(3, 39, length.out = npts),
                       density = dnorm(x, 20.9, 5.6))

SAT.plot <-
  SAT.norm %>%
  ggplot(aes(x, density)) +
  geom_line() + 
  geom_vline(xintercept = 1350, color='red', linetype='dashed') + 
  annotate("text", x = 1400, y = dnorm(1060, 1060, 195), color='red', label='Malaika') +
  labs(x = "Score", y = "Density",  title = "Distribution of SAT scores") +
  theme_classic() + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        plot.title = element_text(size = 10, hjust=0.5))  

ACT.plot <-
  ACT.norm %>%
  ggplot(aes(x, density)) +
  geom_line() + 
  geom_vline(xintercept = 28, color='blue', linetype='dashed') + 
  annotate("text", x = 29, y = dnorm(20.9, 20.9, 5.6), color='blue', label='Jim') +   
  labs(x = "Score", title = "Distribution of ACT scores") +
  theme_classic() + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        plot.title = element_text(size = 10, hjust=0.5))

plot_grid(SAT.plot, ACT.plot, labels = "AUTO")
```


A: Since the scores are measured on different scales we can not directly compare them, however we can measure the difference of each score in terms of units of standard deviation

### Standardized or Z-scores

Differences from the mean, measured in units of standard deviation are called "standardized scores" or "Z-scores"

- The Z score of an observation is the number of standard deviations it falls above or below the mean.
\[
Z_i = \frac{x_i - \mu}{\sigma}
\]


- For our SAT/ACT example above:
\[
\begin{align}
Z_\text{Malaika} &= \frac{1350 - 1060}{195} = 1.49\\
\\
Z_\text{Jim} &= \frac{28 - 20.9}{5.6} = 1.27\\
\end{align}
\]
In this case, Malaika's score is 1.49 standard deviation above the mean, while Jim's is 1.27. Based on this, Malaika scored better than Jim.


```{r, echo=FALSE, fig.width = 6, fig.height = 3.5}
std.norm.plot.labels +
  geom_vline(xintercept = 1.49, color="red", linetype="dashed") + 
  annotate("text", x = 1.75, y = 0.4, color='red', label='Malaika') +
  geom_vline(xintercept = 1.27, color="blue", linetype="dashed") + 
  annotate("text", x = 1.1, y = 0.4, color='blue', label='Jim') +
  labs(x = "Z-scores", title="Comparison of Malaika's and Jim's\n standardized college entrance exam scores")
```


## Standard normal distribution

- If $X \sim N(\mu,\sigma)$ then the standardized distribution, $Z_X \sim N(0,1)$. If $X$ is normally distributed, then the Z-scores based on $X$ have a mean of 0 and a standard deviation of 1. 

- $N(0,1)$ is known as the **standard normal distribution**


## Percentiles

- The *percentile* is the percentage of observations that fall below a given point, $q$

- In R, for a normal distribution the fraction of observations below a given point (the probability that a random observation drawn from the distribution is less than the given value) can be calculatedusing the `pnorm(q, mu, sigma)` function:

    ```{r}
# Malaika's z-score was 1.49. What percentile was she in?
pnorm(1.49) 
```
Therefore, Malaika is approximately at the 93-percentile. Note that we didn't have to include the mean and standard deviation in the call to pnorm because we're dealing with standardized scores, and the defaults for `pnorm` are `mean = 0` and `sd = 1`.  A similar calculation would show that Jim's percentile is `r pnorm(1.27) * 100`. 

```{r, echo=FALSE, fig.width = 6, fig.height = 3.5}
std.norm.plot.labels +
  geom_vline(xintercept = 1.49, color="red", linetype="dashed") + 
  annotate("text", label="Malaika's z-score: 1.49",
           x = 1.49 + 0.1, y = 0.4, color='red', hjust=0) +
  annotate("text", x = 0, y = 0.1, color='black', label='pnorm(1.49) = 0.93',size=5) +
  geom_area(data = filter(std.norm, x < 1.49), alpha=0.25) + 
  labs(x = "Z-scores", title="Malaika scored higher than ~93%\nof other exam takers on the SAT")
```


## Cutoff points

- When we use the `pnorm()` function we give it a given point, $q$, and it gives us the corresponding fraction of values, $p$, that fall below that point in a normal distibution

- If instead we want to specify a fraction $p$, and get the corresponding point, $q$, on the normal distribution, we use the `qnorm(p, mu, sigma)` function.

```{r}
# To get the 75-th percentile (3rd quartile) of SAT scores 
# based on the parameters provided previously
qnorm(0.75, 1060, 195)
```

```{r, echo=FALSE, fig.width = 6, fig.height = 3.5, fig.align = "center", out.width="60%"}
perc.75 <- qnorm(0.75, 1060, 195)

SAT.norm %>%
  ggplot(aes(x, density)) +
  geom_line() + 
  geom_vline(xintercept = perc.75, linetype="dashed") +
  geom_area(data = filter(SAT.norm, x < perc.75), alpha=0.25) +
  annotate("text", x = 1060, y = 0.001, color='black', label='p = 0.75',size=5) +
  annotate("point", x = perc.75, y = 0, color='red', size=4) +
  annotate("text", x = 1300, hjust=0, y = 0.0012, color='red', label='qnorm(0.75, 1060, 195)', size=4) +
  annotate("segment", 
           x = 1295, y = 0.00115, 
           xend = perc.75 + 5, yend = 0 + 0.00005,
           color='red', arrow = arrow(type="closed"), size=1) +
  labs(x = "Score", y = "Density",  title = "Distribution of SAT scores") +
  theme_classic() + 
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        plot.title = element_text(size = 10, hjust=0.5))  
  
```


## 88-95-99.7 Rule

If data are approximately normally distributed:

- ~68% of observations lie within 1 SD of the mean
- ~95% of observations lie within 2 SD of the mean
- ~99.7% of observations lie within 3 SD of the mean


```{r, echo=FALSE, fig.width = 6, fig.height = 3, fig.align = "center", out.width="60%"}

std.norm.plot + 
  geom_area(alpha=0.5) + 
  geom_area(data=filter(std.norm, abs(x) < 2), fill="gray") + 
  geom_area(data=filter(std.norm, abs(x) < 1), fill="oldlace") +
  annotate("segment", x = -1, xend = 1, 
           y = 0.22, yend = 0.22, arrow=arrow(ends="both")) + 
  annotate("segment", x = -2, xend = 2, 
           y = 0.15, yend = 0.15, arrow=arrow(ends="both")) +
  annotate("segment", x = -3, xend = 3, 
           y = 0.08, yend = 0.08, arrow=arrow(ends="both")) +
  annotate("text", x = 0, y = 0.23, label="68%") + 
  annotate("text", x = 0, y = 0.16, label="95%") + 
  annotate("text", x = 0, y = 0.09, label="99.7%")
```


## Assessing normality

### Normal probability plot

```{r, echo=FALSE, warning=FALSE,message=FALSE}
male.heights <- read_tsv("https://github.com/bio304-class/bio304-course-notes/raw/master/datasets/male-heights.txt")
nba <- read_tsv("https://github.com/bio304-class/bio304-course-notes/raw/master/datasets/nba-heights.txt")
```

```{r,echo=FALSE}
mean.height <- mean(male.heights$heights)
sd.height <- sd(male.heights$heights)

height.histo <- 
  male.heights %>%
  ggplot() + 
  geom_histogram(aes(x = heights, y = ..density..),
                 bins=9, alpha=0.5) +
  stat_function(fun = function(x){dnorm(x, mean.height, sd.height)},
                n  = 200) + 
  labs(x = "Height (in)",
       y = "Density")   
  
qq.height <-
  male.heights %>%
  ggplot(aes(sample = heights)) + 
  geom_qq(alpha=0.5, color='steelblue') + 
  geom_qq_line() + 
  labs(x = "Theoretical quantile",
       y = "Height (in)")

plot_grid(height.histo, qq.height)
```




```{r,echo=FALSE}
nba.mean.height <- mean(nba$h.in)
nba.sd.height <- sd(nba$h.in)

nba.height.histo <- 
  nba %>%
  ggplot() + 
  geom_histogram(aes(x = h.in, y = ..density..),
                 breaks=seq(68,92,2), alpha=0.5) +
  stat_function(fun = function(x){dnorm(x, nba.mean.height, nba.sd.height)},
                n  = 200) +
  labs(x = "Height (in)",
       y = "Density")  
  
nba.qq.height <-
  nba %>%
  ggplot(aes(sample = h.in)) + 
  geom_qq(alpha=0.5, color='steelblue') + 
  geom_qq_line() + 
  labs(x = "Theoretical quantile",
       y = "Height (in)")

plot_grid(nba.height.histo, nba.qq.height)
```
